---
title: "dplyr and tidyr"
author: "dzaky jaya"
date: "2022-08-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Manipulate Data

load library

```{r}
library(tidyverse)
```

load data

```{r}
df <- mpg
print(df)

# check null values
any(is.na(df))

# explore number of columns and rows
ncol(df); nrow(df)

# explore structure of a table
str(df)

# explore column names
colnames(df)
```

## select()

function to select column like select clause in SQL

```{r}
# extract specified column
df %>% select(manufacturer, model, year, class)

# extract matched criteria
df %>% select(starts_with(match = "c"))

df %>% select(contains("a"))

# select using index
df %>% select(2:5)

# select last n columns (little tricky)
df %>% select((ncol(df)-2):ncol(df)) # it will extract last 3 columns
```

## rename() and everything()

everything is a function to grab all the remaining unspecified column

```{r}
# rename all column
df1 <- df %>% rename(yes = everything())

# using select and rename simultaneously
df %>% select(mnfc = manufacturer, mod = model, display = displ, everything())
# note that this code doesn't rename original data set, it just rename the displayed data, i.e aliases

#to rename the original data set, use assign
df <- df %>% rename(mnfc = manufacturer, mod = model)
```

## mutate() and transmute()

mutate is a function to create new variable in a data set\
transmute is a function to create new variable in a data set, but it will remove all others variables\

```{r}
# create new variable
df <- df %>% mutate(`avg mile per galon` = (cty+hwy)/2)

df <- df %>% mutate(car = paste(mnfc, mod))

df <- df %>% mutate(`cyl and trans` = paste(cyl, "cylinders and", trans, "transmissions"))
```

## filter() and slice()

filter is a function to filter rows by criteria\
slice is a function to filter rows by its indices

```{r}
# use same order logic as SQL logic
df %>%  select(manufacturer, year, hwy) %>% filter(manufacturer %in% c("audi","ford"))

# if we want to filter manufacturer but not want manufacturer to be displayed, then put the filter prior to select
df %>% filter(manufacturer %in% c("audi","ford")) %>%  select(year, hwy)
# if we want to filter manufacturer but not want manufacturer to be displayed but put the filter after select, it will cause an error

# combining filter and select
df %>%  select(manufacturer, year, hwy) %>% filter(manufacturer %in% c("audi","ford")) %>% group_by(manufacturer, year) %>% summarise(`avg mean` = mean(hwy))
```

slice()

```{r}
# get data from row 20 to 30
df %>% slice(20:30)

# combining slice and select
df %>% select(manufacturer, model, year) %>% slice(20:30)
```

## arrange()

just like `order by` clause in SQL\
default ordering is ascending

```{r}
# order cty in asc and hwy in desc, then display 20 records
print(df %>% arrange(cty, desc(hwy)), n=20)

# using arrange in advance
print(df %>% select(manufacturer, year, hwy) %>% group_by(manufacturer) %>% filter(year == 1999) %>% summarise(`average of hwy` = mean(hwy)) %>% arrange(desc(`average of hwy`)), n=5)
```

## distinct()

just like `distinct` cluase in SQL

```{r}
df %>% distinct(manufacturer, year) %>% arrange(manufacturer, year)
```

## bind\_()

binding either row(s) or column(s) just like rbind and cbind

```{r}
exmp <- data.frame(id = c(1,2,3), names = c("john", "michael", "aries"))

exmp <- bind_rows(exmp, slice(exmp, 2:3))

exmp <- arrange(exmp, id)
```

## sample_n() and sample_frac()

`sample_n()` to sampling number of n rows\
`sample_frac()` to sampling percent of rows

```{r}
# retrieve 20 samples from df without replacement
sample_n(df, 20, replace = F)

# retrieve 20% sample from df
sample_frac(df, 0.2)
```

## summarise()

used with aggregate function
```{r}
# count number of records and number of distinct cty records
df %>% summarise(n(), n_distinct(cty))

# count number of records in advance
df %>% select(manufacturer, model) %>% group_by(manufacturer, model) %>% summarise(numrec = n())
```
the same approach with simpler code line is using `count()`

## count()

```{r}
count(df, manufacturer, model) # will produce exactly same result as in second code line of summarise() part

df %>% group_by(manufacturer) %>% summarise(n())
```

```{r}
df %>% group_by(manufacturer) %>% summarise(`min hwy` = min(hwy), `max hwy` = max(hwy),
                                            `min cyl` = min(cyl), `max cyl` = max(cyl))
```

```{r}
df %>% group_by(manufacturer, model, class, trans) %>% 
        summarise(`mean hwy` = mean(hwy), cars = n()) %>% 
        filter(`mean hwy` > 30) %>% 
        arrange(desc(`mean hwy`)) %>% ungroup()
```

## pivot_longer() and pivot_wider()

transforming table from wide format to long format and vice versa.
```{r}
# create long table example
long <- data.frame(id = 1:6, type = c("a", "b", "b", "c", "b", "a"),
                   total = c(20, 40, 30, 45, 12, 35))

# convert to wide format
wide <- pivot_wider(long, names_from = type, values_from = total)

# bring back wide to long
wide %>% pivot_longer(cols = c("a", "b", "c"), names_to = "type", values_to = "total", values_drop_na = T)

# play in advance
play <- df %>% filter(manufacturer %in% c("jeep", "hyundai", "land rover")) %>% 
        select(model, trans, hwy)

wideplay <- play %>% group_by(model, trans) %>%  summarise(`avg hwy` = mean(hwy)) %>% 
        pivot_wider(names_from = trans, values_from = `avg hwy`)

#convert back with little trick
wideplay %>% pivot_longer(cols = -model, # exclude column model and remain all the rest column
                          names_to = "trans",
                          values_to = "avg hwy",
                          values_drop_na = T)
```

## separate() and unite()

just like split and merge in excel\
`separate`
```{r}
# create dates table just for example
dates <- seq.Date(from = as.Date("2022-01-01"), to = as.Date("2022-12-31"), by = "day")
# place in table format
datedf <- data.frame(date = dates)

sptdatedf <- datedf %>% separate(col = date, into = c("year", "month", "date"), sep = "-") %>% 
        mutate(month = as.numeric(month), 
               date = as.numeric(date)) # remove leading zeros

# other ways to mutate
datedf %>% separate(col = date, into = c("year", "month", "date"), sep = "-") %>% 
        mutate_at(.vars = c("month", "date"), # determine what's columns to mutate
                  .funs = as.numeric) # determine what kind of function to apply
```

`unite`
```{r}
# add leading zeros before merge dates columns
sptdatedf %>% mutate(month = str_pad(width = 2, side = "left", string = month, pad = 0),
                     date = str_pad(width = 2, string = date, side = "left", pad = 0)) %>% 
        unite(col = "full dates", c(year,month,date), sep = "-", remove = F)

# others way to adding leading zeros with mutate_at
sptdatedf %>% mutate_at(.vars = c("month", "date"),
                        .funs = str_pad, width = 2, side = "left", pad = 0) %>% 
        unite(col = "full date", c("year", "month", "date"), sep = "-", remove = F)
```

## pull()

extract column values as a vector
```{r}
df %>% pull(hwy)
```

## combining group_by and mutate

```{r}
df %>% group_by(manufacturer, model) %>% mutate(`avg hwy` = mean(hwy))

# the different between combined group_by-mutate and group_by-summarise is, the first combine
# also added summarised value to column for each corresponding row, while the other one only
# summarise grouped rows without adding it to a column.

df %>% group_by(manufacturer, model) %>% summarise(mean(hwy))
```


## case_when()

just like `case when` in SQL
```{r}
# library data.table is loaded to use like clause like in SQL
library(data.table)

# create a column to identify the transmission type
# whether automatic coded as AT or manual coded as MT
df %>% mutate(`trans type` = case_when(
        trans %like% "auto.*" ~ "AT",
        trans %like% "manual.*" ~ "MT",
        TRUE ~ "NA")
)
```



## mutate FAMILY

`mutate_all()` apply function to all columns\
`mutate_if()` apply function over columns that meet condition\
`mutate_at()` apply function to a specific columns









