---
title: "S1 - assignment"
author: "dzaky jaya"
date: "2022-09-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

load data & library

```{r}
library(tidyverse)

df <- hflights::hflights
class(df)
```

### How many rows and columns are in table hflights?

```{r}
paste("number of rows:", nrow(df)); paste("number of columns:", ncol(df))
```

### How many different carriers are listed in the table (print a table with distinct carrier names)?

```{r}
df %>% count(UniqueCarrier)
```

another way that only display the distinct carriers name

```{r}
df %>% distinct(UniqueCarrier)
```

### Which and how many airports were involved? Consider both origin and destination airports!

```{r}
paste("there are about", nrow(df %>% distinct(Dest)), "dest airport are involved") ; paste("there are about", nrow(df %>% distinct(Origin)), "origin airport are involved")
```

### How many flights were cancelled?

```{r}
df %>% count(Cancelled) %>% filter(Cancelled == 1)

df %>% mutate(status = case_when(Cancelled == 1 ~ "Cancelled",
                                 TRUE ~ "Success")) %>% count(status)
```

# Exercise 2

First, produce a table where statistics for each carrier is shown:\
number of flights per carrier\
total distance flown in miles per carrier\
total actual elapsed time in hours per carrier\
total air time in hours per carrier\
mean distance per flight for each carrier\
mean actual elapsed time in hours per flight for each carrier\
mean air time in hours per flight for each carrier

```{r}
dfcarrier <- df %>% group_by(UniqueCarrier) %>% summarise(numflight = n(),
                                             totdist = sum(Distance),
                                             elaptimehr = sum(ActualElapsedTime, na.rm = T)/60,
                                             airtimehr = sum(AirTime, na.rm = T)/60,
                                             `mean dist/flight` = totdist/numflight,
                                             `mean elaptimehr/flight` = elaptimehr/numflight,
                                             `mean airtimehr/flight` = airtimehr/numflight
                                             ) %>% 
        ungroup()
```

Second, calculate the percentage of total distance flown by top 3 performing carriers VS total
distance flown by remaining carriers. Execute steps:\
 first rank carriers by total distance flown\
 top 3 performers are in one group, remaining carriers are in second group\
 for each group calculate total distance flown\
 for each group calculate %\
```{r}
dfcarrier %>% arrange(desc(totdist)) %>% mutate(rank = row_number(),
                                                group = case_when(rank %in% 1:3 ~ "top3",
                                                                  T ~ "remaining")) %>% 
        group_by(group) %>% summarise(tdgroup = sum(totdist)) %>%
        mutate(percentage = tdgroup/sum(tdgroup)*100) %>% 
        select(group, `percent of total distance` = percentage) %>%
        arrange(desc(`percent of total distance`)) %>% 
        ungroup()
```

# Exercise 3
Modify your main flights table:
 create date column by uniting columns: year, month, day of month
 when uniting columns do not lose source columns (mutate each column - with slightly
different name, before unite operation is executed)
 you will need to parse date column after unite operation
 also you should add leading zeros to month and day of month column before date is
created
 create columns: quarter, week

```{r}
dfmod <- df %>% mutate_at(.vars = c("Month", "DayofMonth"), .funs = str_pad,
                 width = 2, pad = 0, side = "left") %>% 
        unite(col = "Date" ,c(Year, Month, DayofMonth), sep = "-", remove = F) %>% 
        #because we apply str_pad function on Month and DayofMonth, these columns format
        #changed to character, thus we convert it back to integer to simplify the logical code
        mutate(Quarter = case_when(as.integer(Month) %in% 1:3 ~ "Q1",
                                   as.integer(Month) %in% 4:6 ~ "Q2",
                                   as.integer(Month) %in% 7:9 ~ "Q3",
                                   as.integer(Month) %in% 10:12 ~ "Q4"),
               WeekofMonth = case_when(as.integer(DayofMonth) %in% 1:7 ~ "W1",
                                as.integer(DayofMonth) %in% 8:14 ~ "W2",
                                as.integer(DayofMonth) %in% 15:21 ~ "W3",
                                as.integer(DayofMonth) %in% 22:28 ~ "W4",
                                TRUE ~ "W5"))

# convert date column to approriate format
dfmod$Date <- ymd(dfmod$Date)
```

the easiest way to create Date column from Year, Month, and DayOfMonth column while convert it to date format is using lubridate
```{r}
library(lubridate)
dfsimp <- df %>% mutate(Date = make_date(year = Year, month = Month, day = DayofMonth),
              # add column quarter and week
              Quarter = case_when(Month %in% 1:3 ~ "Q1",
                                  Month %in% 4:6 ~ "Q2",
                                  Month %in% 7:9 ~ "Q3",
                                  Month %in% 10:12 ~ "Q4"),
              WeekofMonth = case_when(DayofMonth %in% 1:7 ~ "W1",
                                      DayofMonth %in% 8:14 ~ "W2",
                                      DayofMonth %in% 15:21 ~ "W3",
                                      DayofMonth %in% 22:28 ~ "W4",
                                      TRUE ~ "W5"))
```

Using your modified table try to answer the given questions:
 Is total number of flights increasing or decreasing quarterly?
 Is total distance increasing or decreasing monthly?

```{r}
dfsimp %>% select(Date, Quarter, WeekofMonth) %>%  group_by(Quarter) %>% 
        summarise(`Total Flight` = n()) %>% 
ggplot() + geom_point(aes(x = Quarter, y = `Total Flight`))
```
result : total number of flight was increasing until 3rd quarter, but significantly decreasing after 3rd quarter
```{r}
dfsimp %>% select(Date, Distance) %>% group_by(Month = month(Date, label = T)) %>%
        summarise(`Total Dist` = sum(Distance)) %>% 
        ggplot() + geom_point(aes(x = Month, y = `Total Dist`))
```
result : total number of distance was unable to determine whether increasing or decreasing monthly, because the pattern showed an unstable trend

# Exercise 4
The idea for the last exercise is another data wrangling task, where you will have to use
technique called ”pivoting”. Build a table, that will resemble a heat map by:
 for each carrier and month, calculate total number of flights
 then normalize total number of flights (divide each value with maximum total number of flights, you must get values between 0 and 1!)
 now pivot your table from long to wide format
 so each row is represented with carrier, and each column is represented with month, normalized total number of flights are values in table cells

```{r}
# total number of flight for each carrier and month
carrmo <- dfsimp %>% mutate(Bulan = month(Date, label = T)) %>%
        select(UniqueCarrier, Bulan) %>% count(UniqueCarrier, Bulan) %>% 
        mutate(normtotal = n/max(n)) %>% select(UniqueCarrier, Bulan, normtotal) %>% 
        pivot_wider(names_from = Bulan, values_from = normtotal)

carrmo %>% pivot_longer(cols = c(-"UniqueCarrier"), names_to = "Month", values_to = "Num of Flight") %>% 
        ggplot() + geom_tile(aes(x = Month, y = UniqueCarrier, fill = `Num of Flight`))
```











