---
title: 'data import: readr and tibble'
author: "Dzaky Jaya"
date: "2022-09-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# tibble and triblle

load packages
```{r}
library(tidyverse)
```

tibble
```{r}
tibble("v1" = 1:10,
       "v2" = 10,
       "v3" = v1*v2)
```
tribble
```{r}
tribble(~v1, ~v2, ~v3,
        "x",  2 ,  4 ,
        "y",  5 , 32 ,
        "z",  3 , 23 )
```
tribble make a table like data entry in spreadsheet, while tibble make a table like row wise.

# read_csv and read_csv2

`read_csv` to `,` separated value\
`read_csv2` to `;` separated value

```{r}
# list the files
list.files("./02_05_read_files")
```

# verctor parsing
parse_character
```{r}
parse_character(c("a", "b", "3"))
```
parse_logical
```{r}
parse_logical(c)
```
parse_double
```{r}
parse_double(c("11.2", "13.4"))

# with different decimal mark
parse_double(c("11,5", "13,4"),
             locale = locale(decimal_mark = ","))
```

specify grouping mask
```{r}
parse_number("1,000.24", locale = locale(grouping_mark = ","))
```

# file parsing
```{r}
read_tsv("./02_05_read_files/mpg.tsv",
         col_types = cols(
                 manufacturer = col_character(),
                 model = col_character(),
                 trans = col_character(),
                 drv = col_character(),
                 fl = col_character(),
                 class = col_character(),
                 displ = col_double(),
                 year = col_double(),
                 cyl = col_double(),
                 cty = col_double(),
                 hwy = col_double(),
         ))
```

# other usefull library for import data
`rio`\
`readxl`\
`data.table`
```{r}
install.packages("rio")
install.packages("readxl")
install.packages("data.table")
```

# Exercise 2
In this exercise you will import .csv file called flights_02.csv, which is located in zip file.
When you are importing the file, try to consider the following:\
 assign imported object to R object named df2\
 for importing use function from library readr\
 inside function for importing, define column parsing1\
 maybe you should inspect the .csv file before actual import\
 after the import, check structure of df2 object with str()

```{r}
df2 <- read_csv("./data_import/data_import/flights_02.csv") %>% 
        unite(col = "Date", c(Year, Month, DayofMonth), sep = "-") %>% 
        mutate(Date = lubridate::ymd(Date), FlightNum = as.integer(FlightNum))

# to make date column appear first than the others
df2 <- df2 %>% select(Date, everything())
df2
```

# Exercise 3
Now you will import .csv file called flights 03.csv. When you are importing the file, you
should consider the following:\
 assign imported object to R object named df3\
 for importing use function from library readr\
 inspect the .csv file before actual import\
 function for import should include some additional import strategies (compared to previous example!)\
 HINT: at the point of import: maybe you should parse all columns as characters\
(col types = cols(.default = ”c”))\
 HINT: then inside R you can convert column types and add column names, after the
import operation is executed\
 after the import, check structure of df3 object with str()

```{r}
df3 <- read_delim("./data_import/data_import/flights_03.csv", skip = 12, 
                  col_names = c("UniqueCarrier", "FlightNum", "Date", 
                                "Origin", "Dest", "Distance"),
                  comment = "#", delim = "|", trim_ws = T)

glimpse(df3)
```

# Exercise 4
In the last exercise you will import .csv file called flights 04.csv, which is a larger flat file
with several million rows:\
 import file two times using readr library and data.table’s fread\
 when importing with readr do column parsing at the point of import\
 when importing with fread force all columns to be parsed as characters (colClasses =
”character”)\
 compare execution times for each importing strategy\
\

```{r}
system.time(
        read_csv2("./data_import/data_import/big_table_04.csv", 
                 col_types = cols(.default = "c"))
) ; system.time(
        data.table::fread("./data_import/data_import/big_table_04.csv",
                          colClasses = "character")
)
```